---
title: "Case study: reviewer experience from ADaM metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: reviewer experience from ADaM metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A submission can be technically correct and still be hard to review. Missing
labels, undocumented derivations, thin annotations, and a missing reviewer guide
all slow a reviewer down, and a slow review is a risk. `r4subusability` scores
this reviewer experience directly, on the same evidence footing as quality and
risk.

This case study builds a Define-XML style metadata view from the example CDISC
pilot metadata in `r4subdata`, then measures it.

```{r libs, message = FALSE}
library(r4subusability)
library(r4subdata)
```

## Building a Define-style view

The usability checks read metadata with an `origin` and, for derived variables,
a `derivation`. The example ADaM metadata does not carry those yet, but the trace
mapping does: a direct copy is a predecessor, anything with a real derivation
rule is derived. Joining the two gives a realistic Define-style table.

```{r meta}
meta <- adam_metadata

key <- paste(meta$dataset, meta$variable)
mk  <- paste(trace_mapping$adam_dataset, trace_mapping$adam_var)
idx <- match(key, mk)
has <- !is.na(idx)
deriv <- trace_mapping$derivation_text[idx]

meta$origin <- "Assigned"
meta$origin[has & grepl("Direct copy", deriv)]  <- "Predecessor"
meta$origin[has & !grepl("Direct copy", deriv)] <- "Derived"
meta$derivation <- NA_character_
meta$derivation[has] <- deriv[has]

head(meta[, c("dataset", "variable", "label", "origin", "derivation")], 6)
```

## Measuring usability

`usability_indicators()` runs the four checks and returns evidence rows. The
`assets` argument is the list of submission documents present, which is how the
reviewer guide check knows whether the guide was shipped.

```{r indicators}
ctx <- r4subcore::r4sub_run_context(study_id = "CDISCPILOT01",
                                    environment = "UAT")

ev <- usability_indicators(
  meta,
  assets = c("adrg.pdf", "define.xml"),
  ctx = ctx
)

nrow(ev)
table(ev$indicator_name, ev$result)
```

## Reading the result

`usability_summary()` turns those rows into the reviewer-facing picture: pass
rates per check and an overall usability score.

```{r summary}
usability_summary(ev)
```

The pattern here is typical. Labels are complete and the reviewer guide is
present, but annotation coverage on derived variables is thin. That is a precise,
actionable finding: document the derivations for the variables the check flags,
and the score moves.

## Where it goes

These rows are ordinary R4SUB evidence. They flow into the Usability pillar of
the Submission Confidence Index next to quality, trace, and risk. Reviewer
experience stops being a soft concern raised late and becomes a measured input
to the readiness decision.
