---
title: "Usability Indicators with r4subusability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usability Indicators with r4subusability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r library}
library(r4subusability)
```

## The Usability Pillar

The R4SUB framework organises submission-readiness evidence into four pillars:
quality, traceability, risk, and **usability**. Usability measures how easy it
is for a regulatory reviewer to understand and navigate your submission package
тАФ a dimension that all six major regulatory authority profiles allocate between
15 % and 20 % of the total submission-readiness weight.

`r4subusability` implements the four indicators that make up this pillar:

| ID | Name | What it measures |
|---|---|---|
| U-001 | Variable Label Quality | Label presence, length, and content |
| U-002 | Define-XML Completeness | Origin and derivation documentation |
| U-003 | Annotation Coverage | Proportion of derived variables annotated |
| U-004 | Reviewer Guide Presence | ADRG / SDRG detected in submission package |

Every assessment function returns a validated R4SUB evidence table via
`r4subcore::as_evidence()`, so the output is immediately compatible with the
scoring and profiling layers.

## Configuration: `usability_config_default()`

All four indicators are driven by a shared configuration list. The defaults
reflect the most common regulatory requirements, but you can override any
element:

```{r config}
cfg <- usability_config_default()
str(cfg)
```

Key parameters:

- `label_min_chars` / `label_max_chars` тАФ acceptable label length window
  (default: 3тАУ40 characters).
- `required_origins` тАФ origins that must have a derivation text
  (`"Derived"`, `"Assigned"`).
- `reviewer_guide_keywords` тАФ strings that identify a reviewer guide asset
  (`"ADRG"`, `"SDRG"`, `"reviewers_guide"`, `"reviewer_guide"`, `"ARG"`).
- `weights` тАФ relative indicator weights summing to 1.

```{r config_weights}
cfg$weights
```

## Run Context

All assessment functions require a run context from `r4subcore`. A single
context object is typically created once per pipeline execution:

```{r ctx}
ctx <- r4subcore::r4sub_run_context(study_id = "STUDY01")
```

## U-001: Variable Label Quality (`assess_label_quality()`)

`assess_label_quality()` checks every variable label for:

- **Presence** тАФ a missing or empty label is a `fail` with `severity = "high"`.
- **Minimum length** тАФ a label shorter than `label_min_chars` is a `warn`.
- **Maximum length** тАФ a label longer than `label_max_chars` is a `warn`.
- **Otherwise** тАФ `pass` with `severity = "info"` and `metric_value = 1`.

```{r u001}
meta_labels <- data.frame(
  dataset  = c("ADSL",    "ADSL", "ADAE"),
  variable = c("USUBJID", "AGE",  "AETERM"),
  label    = c(
    "Unique Subject Identifier",  # OK тАФ 26 chars
    "Age",                         # OK тАФ 3 chars (at minimum)
    ""                             # fail тАФ empty label
  ),
  stringsAsFactors = FALSE
)

ev_label <- assess_label_quality(meta_labels, ctx)
ev_label[, c("location", "result", "severity", "metric_value", "message")]
```

## U-002: Define-XML Completeness (`assess_define_completeness()`)

`assess_define_completeness()` verifies that the Define-XML metadata is
populated correctly. For variables whose `origin` is in `required_origins`, a
non-empty `derivation` text is mandatory.

```{r u002}
meta_define <- data.frame(
  dataset    = c("ADSL",    "ADSL",   "ADAE"),
  variable   = c("USUBJID", "AGE",    "AETERM"),
  label      = c(
    "Unique Subject Identifier",
    "Age",
    "Adverse Event Term"
  ),
  origin     = c("CRF",     "Derived", "CRF"),
  derivation = c(NA,        "Derived from BRTHDTC and RFSTDTC", NA),
  codelist   = c(NA,        NA,        NA),
  stringsAsFactors = FALSE
)

ev_define <- assess_define_completeness(meta_define, ctx)
ev_define[, c("location", "result", "severity", "message")]
```

USUBJID and AETERM come from CRF тАФ no derivation required, both pass. AGE is
`"Derived"` and has a derivation text, so it also passes.

A variable with `origin = "Derived"` but no derivation text would fail:

```{r u002_fail}
meta_bad <- data.frame(
  dataset    = "ADSL",
  variable   = "TRTEDT",
  label      = "Treatment End Date",
  origin     = "Derived",
  derivation = NA_character_,
  stringsAsFactors = FALSE
)
ev_bad <- assess_define_completeness(meta_bad, ctx)
ev_bad[, c("location", "result", "severity", "message")]
```

## U-003: Annotation Coverage (`assess_annotation_coverage()`)

`assess_annotation_coverage()` works at the **dataset level**. For each
dataset it computes what proportion of derived variables have non-empty
derivation text:

- `>= 90 %` тЖТ `pass`
- `70 тАУ 89 %` тЖТ `warn`
- `< 70 %` тЖТ `fail`

```{r u003}
meta_annot <- data.frame(
  dataset    = c("ADSL",   "ADSL",   "ADSL",   "ADAE",   "ADAE"),
  variable   = c("AGE",    "SEX",    "TRTEDT", "AESTDTC","AEENDTC"),
  origin     = c("Derived","CRF",    "Derived","CRF",    "CRF"),
  derivation = c(
    "Derived from BRTHDTC and RFSTDTC",
    NA,                                   # CRF тАФ not counted
    NA,                                   # Derived but not documented
    NA,                                   # CRF тАФ not counted
    NA                                    # CRF тАФ not counted
  ),
  stringsAsFactors = FALSE
)

ev_annot <- assess_annotation_coverage(meta_annot, ctx)
ev_annot[, c("asset_id", "result", "metric_value", "message")]
```

ADSL has 2 derived variables, only 1 annotated тЖТ 50 % тЖТ `fail`. ADAE has no
derived variables so coverage is not applicable (`na`).

## U-004: Reviewer Guide Presence (`assess_reviewer_guide()`)

`assess_reviewer_guide()` checks a character vector of asset names for any
keyword that matches `reviewer_guide_keywords` (case-insensitive substring
match). It returns a single evidence row.

```{r u004_pass}
# Submission package includes an ADRG
ev_guide_pass <- assess_reviewer_guide(
  assets = c("ADRG", "define.xml", "datasets"),
  ctx    = ctx
)
ev_guide_pass[, c("result", "severity", "message")]
```

```{r u004_fail}
# Submission package is missing a reviewer guide
ev_guide_fail <- assess_reviewer_guide(
  assets = c("define.xml", "datasets"),
  ctx    = ctx
)
ev_guide_fail[, c("result", "severity", "message")]
```

## Master Function: `usability_indicators()`

`usability_indicators()` runs all four assessments in a single call and returns
one combined evidence table:

```{r master}
meta_full <- data.frame(
  dataset    = c("ADSL",    "ADSL",    "ADAE"),
  variable   = c("USUBJID", "AGE",     "AETERM"),
  label      = c(
    "Unique Subject Identifier",
    "Age",
    "Adverse Event Term"
  ),
  origin     = c("CRF",     "Derived", "CRF"),
  derivation = c(NA, "Derived from BRTHDTC and RFSTDTC", NA),
  stringsAsFactors = FALSE
)

assets <- c("ADRG", "define.xml")

ev_all <- usability_indicators(meta_full, assets = assets, ctx = ctx)

# One row per variable per indicator, plus one row for the reviewer guide
nrow(ev_all)
table(ev_all$indicator_id, ev_all$result)
```

## Summary and Print Method

`usability_summary()` produces a structured `usability_result` object with
per-indicator pass/fail/warn counts and an overall score. The `print` method
renders a colour-coded console report:

```{r summary}
res <- usability_summary(ev_all)
print(res)
```

You can also access the underlying summary data frame programmatically:

```{r summary_df}
res$summary[, c("indicator_id", "indicator_name", "n_pass", "n_fail", "pct_pass")]
```

## Customising the Configuration

Override specific defaults without replacing the entire list:

```{r custom_config, eval = FALSE}
cfg_strict <- usability_config_default()
cfg_strict$label_min_chars <- 5L   # stricter minimum length
cfg_strict$label_max_chars <- 30L  # tighter maximum

ev_strict <- assess_label_quality(meta_full, ctx, config = cfg_strict)
```

## Integration with the R4SUB Ecosystem

The evidence table returned by any `r4subusability` function plugs directly
into the rest of the R4SUB pipeline:

```r
# Downstream тАФ illustrative, requires r4subscore and r4subprofile
score  <- r4subscore::compute_score(ev_all)
report <- r4subprofile::render_profile(score, authority = "FDA")
```

Evidence can also be persisted for audit trails:

```{r persist, eval = FALSE}
tmp <- tempfile(fileext = ".rds")
r4subcore::export_evidence(ev_all, tmp, format = "rds")
ev_reloaded <- r4subcore::import_evidence(tmp, format = "rds")
```
